---
title: "README"
output: md_document
date: "2025-10-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# muffinette - Meta-analysis of networks in microbiome studies

The repository houses the **`muffinette`** R package for multi-study meta-analysis of microbial network properties.

## Dependencies

`muffinette` requires the following `R` package: `devtools` (for installation only). Please install it before installing `muffinette`, which can be done as follows (execute from within a fresh R session):

```{r depends, warning=FALSE, message = FALSE, results = "hide"}
install.packages("devtools")
library(devtools)
```

## Installation

Once the dependencies are installed, `muffinette` can be loaded using the following command:

```{r install, warning=FALSE, message=FALSE, results = "hide"}
devtools::install_github("himelmallick/muffinette", quiet = TRUE)
library(muffinette)
```

## Example Implementation
