---
title: "README"
output: md_document
date: "2026-03-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE)
options(repos = c(CRAN = "https://cloud.r-project.org"))
```

# muffinette - Meta-analysis of networks in microbiome studies

The repository houses the **`muffinette`** R package for multi-study meta-analysis of microbial network properties.

## Dependencies

`muffinette` requires the following `R` package: `devtools` (for installation only). Please install it before installing `muffinette`, which can be done as follows (execute from within a fresh R session):

```{r depends, warning=FALSE, message = FALSE, results = "hide"}
install.packages("devtools")
library(devtools)
```

## Installation

Once the dependencies are installed, `muffinette` can be loaded using the following command:

```{r install, warning=FALSE, message=FALSE, results = "hide"}
devtools::install_github("himelmallick/muffinette", quiet = TRUE)
library(muffinette)
```

## Example Implementation

We showcase the effectiveness of our network-connectivity-based meta-analytic
approach (muffinette) by drawing a comparison with feature-abundance-based
meta-analysis (MMUPHin) based on the CRC data available through the R package
`MMUPHin`. This data consists of  species level relative abundance profiles of
CRC and control patients in the five public studies used in Thomas et al. (2019). The dataset is sourced from the R package `curatedMetagenomicData`.

### Data Pre-processing

```{r load-and-prepare-data}
rm(list=ls())
library(dplyr)
library(MMUPHin)
library(muffinette)
library(vegan)
library(ggplot2)
library(cowplot)
library(ggpubr)


data("CRC_abd", "CRC_meta")

set.seed(2310)

data_meta <- data.frame(sampleID = colnames(CRC_abd),
                        age = CRC_meta$age,
                        gender = as.factor(CRC_meta$gender),
                        response = CRC_meta$study_condition,
                        study = as.factor(CRC_meta$studyID))

data_abd <- CRC_abd # feature-by-sample matrix
rownames(data_meta) <- colnames(data_abd)

filtered_featuretable <- muffinette::filter_abdfeatures(x = t(data_abd),
                                                        abd_threshold = 0,
                                                        prev_threshold = 0.1)
filtered_featuretable <- muffinette::filter_varfeatures(x = filtered_featuretable, topV = NULL)

meta_abd_mat <- t(as.matrix(filtered_featuretable)) ## feature-by-sample matrix
```

### Batch (Study) Effect Correction

```{r batch-correction}
batch_corrected_abd <- MMUPHin::adjust_batch(feature_abd = meta_abd_mat,
                                             batch = "study",
                                             covariates = "response",
                                             data = data_meta)$feature_abd_adj
```


### Visualization of Batch (Study) Effect Correction


```{r visualize-batch-correction}
D_before <- vegan::vegdist(t(meta_abd_mat))
D_after <- vegan::vegdist(t(batch_corrected_abd))

fit_adonis_before <- vegan::adonis2(D_before ~ study, data = data_meta)
fit_adonis_after <- vegan::adonis2(D_after ~ study, data = data_meta)

print(fit_adonis_before)
print(fit_adonis_after)

##############
# Ordination #
##############
# Before
R2_before <- round(fit_adonis_before$R2[1]*100, 1)
pcoa_before <- cmdscale(D_before, eig = TRUE)
ord_before <- as.data.frame(pcoa_before$points)
percent_var_before <- round(pcoa_before$eig / sum(pcoa_before$eig) * 100, 1)[1:2]
before_labels <- c(paste('Axis 1 (', percent_var_before[1], '%)', sep = ''), 
                   paste('Axis 2 (', percent_var_before[2], '%)', sep = ''))

before_phrase <- paste('Before (PERMANOVA R2 = ', R2_before, '%)', sep = '')
colnames(ord_before) <- c('PC1', 'PC2')
ord_before$Study <- data_meta$study

# After
R2_after <- round(fit_adonis_after$R2[1]*100, 1)
pcoa_after <- cmdscale(D_after, eig = TRUE)
ord_after <- as.data.frame(pcoa_after$points)
percent_var_after <- round(pcoa_after$eig / sum(pcoa_after$eig) * 100, 1)[1:2]
after_labels <- c(paste('Axis 1 (', percent_var_after[1], '%)', sep = ''), 
                  paste('Axis 2 (', percent_var_after[2], '%)', sep = ''))

after_phrase <- paste('After (PERMANOVA R2 = ', R2_after, '%)', sep = '')
colnames(ord_after) <- c('PC1', 'PC2')
ord_after$Study <- data_meta$study


# Ordination Plot
p_before <- ggplot(ord_before, aes(x = PC1, y = PC2, color = Study)) + 
    geom_point(size = 4) + 
    theme_bw() + 
    xlab(before_labels[1]) + 
    ylab(before_labels[2]) + 
    ggtitle(before_phrase) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(legend.position ="none")

p_after <- ggplot(ord_after, aes(x = PC1, y = PC2, color = Study)) + 
    geom_point(size = 4) + 
    theme_bw() + 
    xlab(after_labels[1]) + 
    ylab(after_labels[2]) + 
    ggtitle(after_phrase) + 
    theme(plot.title = element_text(hjust = 0.5)) + 
    theme(legend.position ="none")

p <- plot_grid(p_before, p_after, ncol = 2)
p
```


### Meta-analysis of feature abundance data (MMUPHin)


```{r mmuphin}
out_mmuphin <- MMUPHin::lm_meta(feature_abd = batch_corrected_abd,
                            exposure = "response",
                            batch = "study",
                            data = data_meta, control = list(forest_plot = "forest.pdf",
                                                             normalization = 'NONE',
                                                             transform = 'NONE'))


metafits_mmuphin <- out_mmuphin$meta_fits
selected_features_mmuphin <- out_mmuphin$meta_fits[out_mmuphin$meta_fits$qval.fdr < 0.05, 1]
selected_features_mmuphin
```


### Meta-network-analysis (muffinette)


```{r muffinette}
batchvar <- data_meta$study
exposurevar <- data_meta$response

out_muffinette <- muffinette::muffinette(metaAbd = data_abd,
                                batchvar = batchvar, exposurevar = exposurevar,
                                metaData = data_meta,
                                filter = TRUE, abd_threshold = 0, prev_threshold = 0.1, topfeatures = NULL,
                                comp = TRUE, net.est.method = "SparCC",
                                covariates = NULL, ncores = 4, verbose = TRUE, fixseed = 2310, iter = 20, inner_iter = 10, th = 0.1)

metafits_muffinette <- out_muffinette$metafits
selected_features_muffinette <- out_muffinette$metafits[out_muffinette$metafits$qval < 0.05, 1]
selected_features_muffinette
```


### Comparison of the Two Meta-Analysis Approaches


```{r compare, eval=FALSE}
fig_mmuphin <- metafits_mmuphin %>%
  filter(qval.fdr < 0.05) %>%
  mutate(feature = sub("^species:", "", feature)) %>%
  arrange(coef) %>% 
  mutate(feature = factor(feature, levels = feature)) %>%
  ggplot(aes(y = coef, x = feature)) +
  geom_bar(stat = "identity", fill = "lightpink") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 6)) +
  labs(y = NULL, x = NULL,
       title = "MMUPHin")

fig_muffinette <- metafits_muffinette %>%
  filter(qval < 0.05) %>%
  mutate(feature = sub("^species:", "", feature)) %>%
  ggplot(aes(y = coef, x = reorder(feature, coef))) +
  geom_bar(stat = "identity", fill = "lightpink") +
  coord_flip() +
  theme(axis.text.y = element_text(size = 5)) +
  labs(y = NULL, x = NULL, title = "muffinette")

fig <- ggarrange(fig_mmuphin, fig_muffinette, nrow = 2, heights = c(0.75, 1))
fig <- annotate_figure(fig,
                       left = text_grob("Differentially Abundant Species",
                                        size = 8, face = "bold", rot = 90),
                       bottom = text_grob("Estimated Coefficients",
                                          size = 8, hjust = 0.5, face = "bold"))
fig
```


